Stream of Consciousness Literature: A 'Joyceless' Linguistic Landscape

### 2024-04-22 paper/paper.qmd
update results

Thus, the estimand is the correlation between mental health-related words in SOC literature, their frequency and sentiment value as provided by @rTidytext. This is considered in terms of nine selected SOC novels only, namely Joyce's \textit{A Portrait of the Artist as a Young Man and Chamber Music}; Woolf's \textit{Mrs Dalloway and Jacob's Room}; Proust's \textit{Swann Way}; Mansfield's \textit{Bliss and The Garden Party}; and Eliot's \textit{The Waste Land and The Love Song of J. Alfred Prufrock}. Through our analysis, we found that (percentage, number and data here, main results)...

@@

However, I do recognize how decontextualized and reductive this quantification of language can be. When it comes to understand such social and human artifacts as language or literary texts, much is dependent on their contextuality. As such, I will further discuss these weaknesses of the datasets under [Discussion].

@@

The archive depends mainly on volunteer efforts for support, and its literature is chosen and proofread by volunteers. However, the archive doesn't take copyrighted or modern items, even with permission. To submit work to the archive, you need to get copyright, scan or harvest images of book pages, spend hours proofreading and formatting, and ensure the eBook meets Project Gutenberg’s requirements, including being valid HTML, correctly spelled, and compliant  with Project Gutenberg’s requirement. Common points of failures for submitted works include:

@@

In terms of format, Most new Project Gutenberg eBooks are in HTML format. Project Gutenberg will check each work's the HTML validity using the W3C’s online validator. The archive also askss for a plain text version whenever possible. This is because plain text has been around for a long time, is widely accessible on all devices and ensures long-term usability. PDF, Word, and other word processor formats are not used as master formats because they are harder to convert to HTML and update  [@gutenberg].

@@

```{r}
#| echo: false
#| eval: true
#| label: tbl-joyce-token
#| tbl-cap: "Tokenized Novel Text & Word Count (James Joyce)"
#| tbl-subcap: 
#|   - "Tokenized Text"
#|   - "Word Count"
#| layout-ncol: 2
#| warning: false
#| message: false

portrait_clean |>
  rename("Book ID" = gutenberg_id,
         "Book" = book,
         "Author" = author,
         "Word" = word) |>
  tail(3) |>
  kable(
    col.names = c("Book ID", "Book", "Author", "Word"),
    booktabs = TRUE
  ) |>
  kable_styling(latex_options = c("striped", "hold_position"), 
                bootstrap_options = "striped", 
                full_width = FALSE)

@@

Then, to quantify the number of times each word appears in each text, I used `count()` (@tbl-joyce-token). The result manifests into @fig-count. The figure demonstrates that beside the top frequencies of personal names, most of the words are abstract nouns that revolve around the nature of time, and life, e.g. god (appearing 194 times in Joyce's), time (appearing in 4 authors' texts, 121 times for Joyce's, 77 times for Woolf's, 421 times for Proust's, 23 times for Eliot), mind (108 times in Joyce's and 196 times in Proust's), life (appearing 266 times in Proust's and 131 times in Joyce's), etc. This further exemplifies the sense of disassociation, temporality and non-linearity commonly associated to stream-of-consciousness literature and more broadly with mental health themes ([@flow], [@pathology], [@resampling]).

@@

@@

@@

Leveraging setiment analysis from @sentiment

@@

bing_lexicon <- get_sentiments("bing") %>%
  rename(sentiment_binary = sentiment)

@@

sentiment_category_binary <- sentiment_combined |>
  group_by(author, sentiment_binary) |>
  count() |>
  group_by(author) |>
  mutate(percentage = n / sum(n) * 100)

# Plot sentiment values
ggplot(sentiment_category_binary, aes(x = author, fill = sentiment_binary, y = percentage)) +
  geom_bar(stat = "identity") +
  labs(
       x = "Author", y = "Percentage") +
  scale_fill_manual(values = c("positive" = "gray80", "negative" = "gray20", "neutral" = "gray50")) + 
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

sentiment_category <- sentiment_combined |>
  group_by(author, sentiment) |>
  count() |>
  group_by(author) |>
  mutate(percentage = n / sum(n) * 100)  |>
  group_by(author) |>
  arrange(desc(percentage))

# Plot sentiment values
ggplot(sentiment_category, aes(x = author, fill = sentiment, y = percentage)) +
  geom_bar(stat = "identity") +
  labs(
       x = "Author", y = "Percentage") +
  scale_fill_manual(values = c("negative" = "gray1", "sadness" = "gray10", "anger" = "gray20", "disgust" = "gray30", "fear" = "gray40", "joy" = "gray50", "trust" = "gray60", "anticipation" = "gray70", "positive" = "gray80", "surprise" = "gray90")) + 
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

@@

@@

sentiment_value <- sentiment_combined |>
  inner_join(get_sentiments("bing")) |>
  group_by(author, value, sentiment) |>
  count() |>
  group_by(author) |>
  mutate(percentage = n / sum(n) * 100)

ggplot(sentiment_value, aes(x = value, y = percentage, color = sentiment, group = sentiment)) +
  geom_line() +
  facet_wrap(~ author, scales = "free") +
  labs(
       x = "Sentiment Value", y = "Percentage") +
  scale_color_manual(values = c("positive" = "black", "negative" = "gray")) + # Adjust line colors
  theme_minimal()
```

@@

## Combined Texts: Trends, Word Networks, Bigram and Trigram Analsis

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false

combined_text|>
  generate_word_graph(minimum_n = 400, minimum_correlation = 1)
```
# Discussion

@@

@@

While these datasets aim to illuminate on the general time trends of childlessness with each variable measured, they are often de-contextualized for the same purpose. For instance, in Table 6, there is no clarifying note on whether people who declared themselves as ‘divorced’ are divorced once or more. This can be extremely narrowing, and exclude nuances when analyzing the contexts behind childlessness.
### Uneven Novel Length and Categorization of Authors

@@

### Project Gutenberg's Focus on the Canon

Along the search for other stream of consciousness authors to include in the generated datasets, only a few-those that are in the SOC canon, mostly white authors-are present in Project Gutenberg, without including lesser known SOC authors, which could have steared the data analysis, still, towards their preconception: being rueful and angst-ridden

>>>>

Thus, the estimand is the correlation between mental health-related themes and words in SOC literature, their frequency and sentiment value as provided by @rTidytext. This is considered in terms of nine selected SOC novels only, namely Joyce's \textit{A Portrait of the Artist as a Young Man and Chamber Music}; Woolf's \textit{Mrs Dalloway and Jacob's Room}; Proust's \textit{Swann Way}; Mansfield's \textit{Bliss and The Garden Party}; and Eliot's \textit{The Waste Land and The Love Song of J. Alfred Prufrock}. Through our analysis, we found that (percentage, number and data here, main results)...

@@

However, I do recognize how reductive this quantification of language can be. When it comes to understand such social and human artifacts as language or literary texts, much is dependent on their context. As such, I will further discuss these weaknesses of the datasets under [Discussion].

@@

The archive relies heavily on volunteers for support and content selection, but it does not accept copyrighted or modern works, even with permission. To submit work, individuals must obtain copyright clearance, scan or capture book images, dedicate hours to proofreading and formatting, and ensure compliance with Project Gutenberg’s guidelines, including valid HTML and correct spelling. Common points of failures for submitted works include:

@@

Most new Project Gutenberg eBooks are in HTML format, which is validated using the W3C’s online validator. Additionally, the archive prefers plain text versions whenever possible due to their universal accessibility, and long-term usability. PDF, Word, and other word processor formats are avoided as master formats due to the challenges involved in converting them to HTML and updating them [@gutenberg].

@@

```{r}
#| echo: false
#| eval: true
#| label: tbl-joyce-token
#| tbl-cap: Tokenized Novel Text & Word Count after Tokenization (James Joyce)
#| tbl-subcap: ["Tokenized Text", "Word Count"]
#| layout-ncol: 2
#| warning: false
#| message: false

portrait_clean <- portrait_clean |>
  select(-book)

portrait_clean |>
  rename("Book ID" = gutenberg_id,
         "Author" = author,
         "Word" = word) |>
  tail(3) |>
  kable(
    col.names = c("Book ID", "Author", "Word"),
    booktabs = TRUE
  ) |>
  kable_styling(latex_options = c("striped", "hold_position"), 
                bootstrap_options = "striped", 
                full_width = FALSE)

@@

Then, to quantify the number of times each word appears in each text, I used `count()` (@tbl-joyce-token). The result manifests into @fig-count. The figure demonstrates that beside the top frequencies of personal names, most of the words are abstract nouns that revolve around the nature of time, and life, e.g. god (appearing 194 times in Joyce's), time (appearing in 4 authors' texts, 121 times for Joyce's, 77 times for Woolf's, 421 times for Proust's, 23 times for Eliot), mind (108 times in Joyce's and 196 times in Proust's), life (appearing 266 times in Proust's and 131 times in Joyce's), etc. This further exemplifies the sense of disassociation, temporality and non-linearity commonly associated to stream-of-consciousness literature and more broadly with mental health themes [@flow] [@pathology] [@resampling].

@@

#reshape the frequency datasets and add sentiment values
bing_lexicon <- get_sentiments("bing") %>%
  rename(sentiment_binary = sentiment)

@@

\newpage
```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false

combined_text|>
  generate_word_graph(minimum_n = 400, minimum_correlation = 1)
```

@@

To further understand the emotional landscape and mental health themes related to SOC literature, I've combined all nine SOC texts to conduct an oveall sentiment analysis [@sentiment].

For Bing index, @fig-sentiment-category shows that, SOC authors' literary works are predominantly negative. All authors, except for Mansfield, demonstrates this. 68% of Joyce's works are negative compared to his 32% positive, similar to Eliot's, which is also the highest statistic out of all five authors. Woolf comes second with 60% words carrying negative meanings, while 40% are negative. Proust comes fourth with  57% negative and 43% positive. While Mansfield has words with positive connotations, at 51%, this isn't much different from her 49% negative words.

The NRC index in @fig-sentiment-category further details the emotional nuances of SOC literature. The dominant emotional connotation in SOC literature's vocabulary is negative, standing at 16% to 22%. While the second highest percentage range belong to words with positive connotation, ranging from 11% to 17%, most of the other bars are occupied by words with negative-leaning emotions, such as sadness, fear, and anger. Sadness-related words, in particular, ranges the third highest, from 11% to 15%.

@@

bing_lexicon <- get_sentiments("bing") |>
  rename(sentiment_binary = sentiment)

@@

sentiment_category_binary <- sentiment_combined |>
  group_by(author, sentiment_binary) |>
  count() |>
  group_by(author) |>
  mutate(percentage = round(n / sum(n) * 100))

# Plot sentiment values
ggplot(sentiment_category_binary, aes(x = author, fill = sentiment_binary, y = percentage)) +
  geom_bar(stat = "identity") +
    geom_text(aes(label = percentage),position = position_stack(vjust = 0.5), color = "white", size = 3) +
  labs(
       x = "Author", y = "Percentage", fill = "Sentiment") +
  scale_fill_manual(values = c("positive" = "gray80", "negative" = "gray20", "neutral" = "gray50")) + 
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),  legend.key.size = unit(0.5, "cm"))

sentiment_category <- sentiment_combined |>
  group_by(author, sentiment) |>
  count() |>
  group_by(author) |>
  mutate(percentage = round(n / sum(n) * 100))  |>
  group_by(author) |>
  arrange(desc(percentage))

# Plot sentiment values
ggplot(sentiment_category, aes(x = author, fill = sentiment, y = percentage)) +
  geom_bar(stat = "identity") +
   geom_text(aes(label = percentage),  position = position_stack(vjust = 0.5), color = "white", size = 3) +
  labs(
       x = "Author", y = "Percentage", fill = "Sentiment") +
  scale_fill_manual(values = c("negative" = "gray1", "sadness" = "gray10", "anger" = "gray20", "disgust" = "gray30", "fear" = "gray40", "joy" = "gray50", "trust" = "gray60", "anticipation" = "gray70", "positive" = "gray80", "surprise" = "gray90")) + 
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
       legend.key.size = unit(0.5, "cm"))

@@

The AFINN index in @fig-sentiment-number portrays a similar trend with its more quantitative sentiment analysis. On a scale of -5 to 5, SOC literature from these five authors mostly carry a negative sentiment, observable in the significant difference in height between the negative trendlines and the positive trendlines. Eliot's works are the most negative with 37% of words being -2 on the emotional scale between -5 and 5. Respectively, Woolf, Proust and Joyce tail beind at 33%, 32% and 31%, averaging out to be ~-2.2 on the emotional scale between -5 and 5. While most a large percentage Mansfield's works are positive, her highest percentage is negative on the AFINN index, at 24% for a scale of about -2.2 between -5 and 5.

This further suggests that the SOC literary landscape is inundated with negative, disgust, fear and sadness-leaning vocabulary. These trends reflect the incoherent mental spaces, depressive perceptions, and existential angst inherent to SOC literature. These are the central themes often conveyed by SOC authors through their exploration of the unconscious mind and repressed emotions, reflecting the societal concerns of the late 19th to mid-20th century West [@pathology] [@resampling].

@@

sentiment_value <- sentiment_combined |>
  inner_join(get_sentiments("bing")) |>
  group_by(author, value, sentiment) |>
  count() |>
  group_by(author) |>
  mutate(percentage = round(n / sum(n) * 100))

ggplot(sentiment_value, aes(x = value, y = percentage, color = sentiment, group = sentiment)) +
  geom_line() +
    geom_text(aes(label = percentage),position = position_stack(vjust = 0.5), color = "black", size = 2) +
  facet_wrap(~ author, scales = "free") +
  labs(
       x = "Sentiment Value", y = "Percentage") +
  scale_color_manual(values = c("positive" = "black", "negative" = "gray")) + # Adjust line colors
  theme_minimal()
```

@@

# Discussion

@@

Stream of consciousness literature emerged as a significant literary technique in the early 20th century, particularly associated with the modernist movement in literature. Several factors contributed to its rise and prominence:

Psychological Realism: One of the primary motivations for the development of stream of consciousness was the desire to represent the inner workings of the human mind more accurately. Writers sought to capture the complexity of thought processes, emotions, and perceptions in a more realistic and nuanced manner.
Response to Realism and Naturalism: Stream of consciousness literature can be seen as a reaction against the more structured and external focus of realism and naturalism. Instead of portraying characters and events objectively, authors wanted to delve deeper into subjective experiences and individual consciousness.
Technological and Cultural Changes: The early 20th century was a period of rapid technological advancement and social change, including the rise of psychology as a discipline. These developments influenced writers to explore new narrative techniques and modes of representation.
Literary Experimentation: Modernist writers were often interested in pushing the boundaries of traditional narrative forms and experimenting with language, style, and structure. Stream of consciousness offered a way to challenge conventional storytelling techniques and engage readers in more innovative ways.
Exploration of Identity and Perception: Stream of consciousness literature allowed writers to explore themes related to identity, consciousness, and perception. By immersing readers in the internal thoughts and sensations of characters, authors could interrogate questions about the nature of reality, self-awareness, and the construction of meaning.
Overall, stream of consciousness literature became a prominent feature of modernist writing due to its capacity to convey the complexity of human experience and its ability to reflect the uncertainties and ambiguities of the modern world.

The rise of modernism in literature, art, and culture during the late 19th and early 20th centuries was influenced by several interconnected factors:

Industrialization and Urbanization: The rapid industrialization and urbanization of society brought about significant changes in the way people lived, worked, and interacted. This upheaval in traditional social structures and lifestyles led to a sense of dislocation and fragmentation, which modernist artists and writers sought to capture in their work.
Technological Advances: Advances in technology, particularly in transportation, communication, and mass media, reshaped human experience and perception of time and space. Innovations such as the telegraph, telephone, and photography altered how people communicated, experienced distance, and understood reality.
Crisis of Faith and Meaning: The late 19th and early 20th centuries witnessed a crisis of faith and meaning, as traditional religious and philosophical beliefs came into question. Developments such as Darwin's theory of evolution, Freudian psychology, and existential philosophy challenged established notions of human existence, morality, and purpose.
World Wars and Political Turmoil: The turmoil of World War I and the interwar period brought about profound changes in society, politics, and culture. The devastation of war, the collapse of empires, and the emergence of new ideologies (such as communism and fascism) shattered old certainties and created an atmosphere of disillusionment and uncertainty.
Intellectual and Cultural Exchange: The modernist movement was characterized by a spirit of intellectual and cultural exchange, with artists and writers drawing inspiration from diverse sources, including non-Western cultures, folk traditions, and avant-garde movements. This openness to new ideas and influences fostered experimentation and innovation in the arts.
Quest for Individuality and Authenticity: Modernism rejected the conventions and constraints of traditional artistic forms and sought to express individual subjectivity and experience in new and unconventional ways. Artists and writers embraced themes of alienation, identity, and the search for authenticity in an increasingly fragmented and alienating world.

@@

While these datasets aim to illuminate on the general time trends of childlessness with each variable measured, they are often de-contextualized for the same purpose. For instance, in Table 6, there is no clarifying note on whether people who declared themselves as ‘divorced’ are divorced once or more. This can be extremely narrowing, and exclude nuances when analyzing the contexts behind childlessness.

### Reliance on Three Different Indices

@@
